Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{{DISTRIBUTION STATEMENT A. Approved for public release.} }

### 2021-05-14 paper/synthesis.tex
First pass of monitor generation

A filter is intended to be very simple; it is expected to have one
input port and one output; messages on the input that the filter
policy admits pass unchanged to the output port; all others are
dropped (not passed on). We have explored in our work two kinds of
filter. In the first, a relatively shallow scan of the input buffer
can enforce the policy. For example, we have used the expressive power
of Contiguity Types \cite{contiguity-types} to enforce
\emph{lightweight} bounds constraints on GPS coordinates in UxAS
messages. On the other hand, the second kind of filter will parse the
input buffer into a data structure specified in AGREE and apply an
user-defined \emph{well-formedness} property, also specified in AGREE,
to the data. This allows arbitrarily complex well-formedness checking.

The decision of a filter is made and performed within one thread
invocation. Thus, in its given time slice, the filter does the following:

@@

\item the wellformedness predicate is evaluated,

@@

\item the filter yields control
\end{enumerate}

@@

The role of partiality should be emphasized: steps 2 and 3 can fail;
the data might not be parseable or the wellformedness computation
could be badly written and fail at run time. In such cases, the filter
should recover and yield control without passing the input onwards. In
these cases, the filter is behaving as it should, but there are also
cases in which a correctly specified filter would not meet its
specification at runtime. This situation arises when the
filter \emph{ought} to accept a message, but lack of resources result
in the filter failing to do so. Examples of this would be, for
example, if the parse of a message needed more space than allocated;
another example would be if the timeslice provided by the scheduler is
too small for the wellformedness computation to finish.

@@

{\emph{Need some discussion of filters and their step-wise properties
  in relation to their infinitary properties. Reference to Johannes'
  work.}

\newsavebox{\contig}
\begin{lrbox}{\contig}
\begin{lstlisting}[style=myML]
  Waypoint =
    {Latitude  : f64
     Longitude : f64
     Altitude  : f32
     Check     : Assert
      (~90.0 <= Latitude and Latitude <= 90.0 and
       ~180.0 <= Longitude and Longitude <= 180.0 and
       1000.0 <= Altitude and Altitude <= 15000.0)}

  AutomationResponse =
    {TaskID : i64
     Length : u8
     Waypoints : Waypoint [3]}
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\contig}}
    \end{tabular}
  \end{center}
  \caption{Contiguity type specification for filter.}
  \label{fig:filter-contig}
\end{figure}

@@

The contiguity type specification for the filter is shown in
\figref{fig:filter-contig}. The synthesized CakeML code for the filter is shown in
\figref{fig:filter-cakeml}. The code is called at dispatch by the
scheduler. The \texttt{API.callFFI} is the link to the communication
fabric to capture input and provide output. The body of the function
restates the filter contract to make the appropriate assignments in a
way that matches the truth value of the predicate in the filter
guarantee.  The auto-generated AGREE specification raises an alert
output when the relation is violated.

@@

Monitors, since they are intended to track and analyze the externally
visible behavior of system components through time, require more
computational features than filters. In particular, our conception of
a monitor is a predicate on its input (and output) streams, being able
to access the value of a stream at any earlier point in time, if
necessary. Thus monitors commonly use state to keep track of earlier
values, unlike filters which are stateless. This reasoning leads us to
specify the computation for a monitor component by a step function of
the following form:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
Thus, in its given timeslice, a monitor evaluates the \konst{stepFn} on its inputs and the current values of the state variables. In detail, it takes the following steps:

@@

\item each available input is parsed into data of the type specified
by the port type;

\item the stateful variables in $\mathit{stateVars}$  are evaluated in dependency order.

\item values of outputs are computed

\item outputs are written and the new state is written

\item control is yielded
\end{enumerate}

Our earlier remarks on partiality apply here too, of course.

The scheduler \emph{activates} components in some order.It is an
obligation on the system that the scheduler follows some sensible
partial order of component activation and allows each component
sufficient time for its computation.  Activating a monitor component
takes the form of the following pseudo-code:
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stepFn} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\paragraph{Initialization}
A monitor may need to accumulate a certain minimum number of
observations before being able to make a meaningful assessment of
behavior. Until that threshold is attained, the monitor is essentially
in a kind of \emph{initialization} phase. In order for correct code to
be generated, monitor specifications need to spell out the values of
output ports when in an initialization phase. For example, suppose a
monitor does some kind of differential assessment of inputs at
adjacent timeslices, alerting when (say) the measured location of a
UAV at times $t$ and $t+1$ is such that the distance between the two
locations is unusually large. Such a monitor needs two measurements
before making its first judgement, but at the time of its first
output, only one measurement will have been made. The specification
for must explicitly state what the correct first output is.

>>>>

A filter is intended to be simple, although it may make deep semantic
checks. A filter has one input port and one output; messages on the
input that the filter policy admits pass unchanged to the output port;
all others are dropped (not passed on). We have investigated two kinds
of filter. In the first, a relatively shallow scan of the input
suffices to enforce the policy. For example, we have used the
expressive power of Contiguity Types \cite{contiguity-types} to
enforce \emph{lightweight} bounds constraints on GPS coordinates in
UxAS messages. On the other hand, a filter may need to parse the input
buffer into a data structure specified in AGREE and apply a
user-defined \emph{well-formedness} property, also specified in AGREE,
to the data. Arbitrarily complex well-formedness checks can be made in
this way. \figref{fig:filter-spec} shows a combination where the
checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+}
depends on an underlying check specified by the contiguity type
checking bounds on waypoints.

The verdict of a filter is made and performed within one thread
invocation. Thus, in its given time slice, the filter must complete the
following steps:

@@

\item the wellformedness predicate is evaluated on the input,

@@

\item the filter yields control.
\end{enumerate}

@@

Partiality is an important consideration: steps 2 and 3 above can
fail; the data might not be parseable or the wellformedness
computation could be badly written and fail at runtime. In such cases,
the filter should recover and yield control without passing the input
onwards. In these cases, the filter is behaving as it should, but we
must also guard against situations in which a \emph{correctly
  specified} filter fails at runtime. This kind of defect arises when
the filter \emph{ought} to accept a message, but lack of resources
results in the filter failing to do so. For example, the parse of a
message might need more space than has been allocated; another example
could be if the timeslice provided by the scheduler is too short for
the wellformedness computation to finish. Thus resource bounds need to
be included in the correctness argument.

@@

%% \emph{Need some discussion of filters and their step-wise properties
%%   in relation to their infinitary properties. Reference to Johannes'
%%   work.}

\newsavebox{\contig}
\begin{lrbox}{\contig}
\begin{lstlisting}[style=myML]
  Waypoint =
    {Latitude  : f64
     Longitude : f64
     Altitude  : f32
     Check     : Assert
      (~90.0 <= Latitude and Latitude <= 90.0 andp
       ~180.0 <= Longitude and Longitude <= 180.0 and
       1000.0 <= Altitude and Altitude <= 15000.0)}

  AutomationResponse =
    {TaskID : i64
     Length : u8
     Waypoints : Waypoint [3]}

 fun WELL_FORMED_AUTOMATION_RESPONSE(aresp) =
   (forall wpt in aresp.Waypoints, WELL_FORMED_WAYPOINT(wpt))
   and ... ;
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\contig}}
    \end{tabular}
  \end{center}
  \caption{Filter specification.}
  \label{fig:filter-spec}
\end{figure}

@@

The contiguity type specification and well-formedness predicate for
the filter are shown in \figref{fig:filter-spec} and the synthesized
CakeML code is in \figref{fig:filter-cakeml}. The code is called at
dispatch by the scheduler. The \texttt{API.callFFI} is the link to the
communication fabric to capture input and provide output. The body of
the function restates the filter contract to make the appropriate
assignments in a way that matches the truth value of the predicate in
the filter guarantee.  The auto-generated AGREE specification raises
an alert output when the relation is violated.

@@

Monitors are intended to track and analyze the externally visible
behavior of system components through time. Therefore, they require
more extensive computational ability than filters. In particular, our
basic notion of a monitor is that it embodies a predicate over its
input and output streams, able to access the value of a stream at any
earlier point in time, if necessary. Monitors commonly use state to
keep track of earlier values, unlike filters which, for us, are
typically stateless components. (However, there is nothing in our
approach that forbids stateful filters: they can be realized by
monitors.) A monitor specification is mapped by code generation to a
state transformation function of the following abstract type:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]

The system scheduler \emph{activates} components in some order. It is
an obligation on the system that the scheduler follows some sensible
partial order of component activation and allows each component
sufficient time for its computation.  Activating a monitor component
takes the form of the following pseudo-code, in which the monitor
evaluates the \konst{stepFn} on its current inputs and the current
values of the state variables, returning the new state and the output
values.
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stepFn} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\subsubsection{Initialization}

A monitor may need to accumulate a certain minimum number of
observations before being able to make a meaningful assessment of
behavior. Until that threshold is attained, the monitor is essentially
in a kind of \emph{initialization} phase. In order for correct code to
be generated, monitor specifications need to spell out the values of
output ports when in an initialization phase. For example, suppose a
monitor does some kind of differential assessment of inputs at
adjacent timeslices, alerting when (say) the measured location of a
UAV at times $t$ and $t+1$ is such that the distance between the two
locations is unusually large. Such a monitor needs two measurements
before making its first judgement, but at the time of its first
output, only one measurement will have been made. The specification
must then explicitly state what the correct first output is.

\subsubsection{Step function}

The \konst{stepFn} works as follows:

@@

\item Each input is parsed into data of the type specified by the port
  type.

\item New values for the state variables are computed, in dependency
  order. The discussion above on initialization now comes into
  play. Suppose the variable declarations have the following form:
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
In the generated code, for the first invocation of \konst{stepFn} only,
the initializations are executed in order:
\[
\begin{array}{l}
  v_1 = i_1; \\
  \cdots \\
  v_n = i_n; \\
\end{array}
\]
In all subsequent steps, the \emph{non-initialization} assignments are performed:
\[
\begin{array}{l}
  v_1 = e_1; \\
  \cdots \\
  v_n = e_n; \\
\end{array}
\]

\item Values of the outputs are computed;

\item Outputs are written and the new state is written;

\item Control is yielded.
\end{enumerate}

The \konst{stepFn} for the monitor of the example is displayed in \figref{fig:monitor-cakeml}.

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
stepFn (Request,Response)
       (req,rsp,current,previous,policy,alert) =
let val stateVars' =
     if !initStep then
        let val req = event(Request)
            val rsp = event(Response)
            val current = (req = rsp)
            val previous = req and not(rsp)
            val policy = current or previous
            alert = not policy
        in (req,rsp,current,previous,policy,alert)
        end
     else
        let val req = event(Request)
            val rsp = event(Response)
            val current = (req = rsp)
            val previous = pre(req and not rsp) and (not req and rsp)
            val policy = current or previous
            alert = (is_latched and pre(alert)) or not(policy)
        in (req,rsp,current,previous,policy,alert)
        end
    val (_,rsp',_,_,_,alert') = stateVars'
    val Alert = if alert' then Some () else None
    val Output =
       if alert' then None else
       if rsp'   then Some Response
       else None
in
   (stateVars', (Alert,Output))
end
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the filter.}
  \label{fig:monitor-cakeml}
\end{figure}

\subsection{Component Behavior}

Possibly some material to add here ...
